---
title: "Untitled"
output: html_document
date: "2024-06-04"
---

#Objectives

#SETUP
```{r}
library(tidyverse)
library(cowplot)
library(ConnMatTools)
library(vctrs)
library(fitdistrplus)
library(boot)
library(MASS)
```

#INs and OUTs
```{r}
#ins
#Re values in
re.in<-"../cluster_linelist/2024-08-22_Results/Re.epochal.clusters.csv"

#seed size in
seed.in<-"../cluster_linelist/2024-08-22_Results/Size2017.clusters.csv"

#clust list in 
clust.comp.in<-"../cluster_composition/2024-05-30_Results/ClusterComposition.csv"

linelist.in<-"../cluster_linelist/2024-08-22_Results/cluster.linelist.csv"

#outs
today<-Sys.Date()
f.out<-paste0(today,"_SBPresults/")
if(!dir.exists(f.out)){dir.create(f.out)}
```

#plot theme
```{r}
pubTheme<-theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                panel.background=element_rect("grey95"), 
                axis.line = element_line(colour = "black"),
                text=element_text(size=10,face="bold"))

riskcolz<-c("darkred","darkblue","darkorange3","darkgreen")
names(riskcolz)<-c("All","gbMSM","PWID","HET")

prepcolz<-c("maroon","orange3")
names(prepcolz)<-c("PrEP","No PrEP")
```

# SETUP extra data
```{r}
#Re epochal predicted and observed
re.t<-read.csv(re.in)
re.t$date_start<-as.Date(re.t$date_start)
re.t$date_end<-as.Date(re.t$date_end)

#size at the end of 2017
seed.df<-read.csv(seed.in)
seed.df<-seed.df[,-1]
#in each simulation, from the size in 2017, draw probability not on ART and not virally suppressed 
#multiply by seed to get potential infectious seed size

#clusters that are active and at least med size
clust.comp<-read.csv(clust.comp.in)
clust.comp<-clust.comp %>% filter (active==TRUE) %>% filter(medium==TRUE )
clust.unq<-clust.comp$clusterID #ACTIVE clusters only
ncl<-length(clust.unq)
#add a risk col

# find primary risk
clust.comp$risk<-NA
for (i in 1:nrow(clust.comp)){
  #If NA, just say mixed
  if(is.na(clust.comp$perc_MSM[i])){
    clust.comp$risk[i]<-"mixed";next}

  if(clust.comp$perc_MSM[i]>=50){
    clust.comp$risk[i]<-"gbMSM"; next
  }
  if(clust.comp$perc_IDU[i]>=50){
    clust.comp$risk[i]<-"PWID"; next
  }
  if(clust.comp$perc_HET[i]>=50){
    clust.comp$risk[i]<-"HET"; next
  }
  #if none of these, then mixed
  clust.comp$risk[i]<-"mixed"
}


clust.comp$active3<-FALSE
clust.comp$active3[which(clust.comp$n_seroconv_2018>=3)]<-TRUE
med.active.clust<-clust.comp$clusterID[which(clust.comp$active3==TRUE & clust.comp$medium==TRUE)]# >=size 10
active.clust<-clust.comp$clusterID[which(clust.comp$active==TRUE)]
#filter on this later
# all(clust.unq %in% re.epoch$clusterID)
# all(clust.unq %in% seed.df$clusterID)

#clusters with no size in 2017 didn't start until 2018 or later
#assign them size 1
nosz<-clust.unq[which(!clust.unq %in% seed.df$clusterID)]
seed.df2<-seed.df[1:length(nosz),]
seed.df2$clusterID<-nosz
seed.df2$size2017<-1
seed.df<-seed.df %>% bind_rows(seed.df2)
# all(clust.unq %in% seed.df$clusterID)

## Observed new cases by cluster
#Observed values
#observed line lists of cluster diagnoses through time
clust.linelist<-read.csv(linelist.in)

#starts on
dayzero<-as.Date("2018-01-01")

#for each cluster, compare cumulative observed to predicted cases during date range
clust.obs.cum<-clust.linelist %>% group_by(clusterID) %>%
  # filter(clusterID %in% med.active.clust) %>%
  filter(FIRST_VL_DATE>=dayzero) %>%
  # group_by(FIRST_VL_DATE) %>%
  dplyr::summarize(clusterID=clusterID[1],
                   total.observed=n())
  
#compare daily observed and predicted new members
clust.obs.cum$type<-"Observed"
clust.obs.cum.medact<-clust.obs.cum %>%
  filter(clusterID %in% med.active.clust) 

```

# Sticky  (and stochastic) branching process SBP
```{r}
# Inspired by code from
# J. Riou, C. L. Althaus, Pattern of early human-to-human transmission of Wuhan 2019 novel coronavirus (2019-nCoV), December 2019 to January 2020. Eurosurveillance 25, 2000058 (2020).
# C. L. Althaus, Ebola superspreading. Lancet Infect. Dis. 15, 507–508 (2015).
# Premise of NEGBIN offspring r0 k from Lloyd-Smith et al. 2005
# and from our SC2 travel restrictions in canada paper

# simulation conditions 

#Run with observed or predicted Re values
#first run Re observerd intial=T, then Re observed inital=F, then Re_noPrEP with initial=f
type.re<-"Re_NoPrEP"#"Re_observed" # #  #
initialrun<-FALSE #initial run TRUE if calculating seed adjust

#after 2018
re.epoch<-re.t %>% filter(date_start >= as.Date("2018-01-01")) %>%
  filter(Type==type.re)

#if no value
default.re<-0.8

# Number of simulation runs per cluster
n.runs<-4000 

#serial interval (gamma dis)
si.mu<-365
si.sd<-182

#calculate scale, shape and rate for SI
#each run: walk around with rlnorm or unif small variation
si.z<-gammaParamsConvert(mean=si.mu, sd=si.sd)
si.shape<-si.z$shape
si.scale<-si.z$scale
si.rate<-1/si.scale

# k dispersion parameter
#each run runif()
k.mean<-0.2
k.min<-0.1
k.max<-0.3

# max time to run simulation
max.long<-as.numeric(diff.Date(as.Date(c("2018-01-01","2022-12-31")))) #days

#progress to 95-95-95 targets (Lima et al)
p.caseperinfect<-p.diagnosed<-0.82 #P(diagnosed|infected) 
p.art<-0.76 #P(on ART|diagnosed)
p.suppressed<-0.83 #P(viral suppressed|on ART)
p.seqpercase<-0.95 #P(sequence|diagnosed)

#mutate initial seed size based on p.art p.suppressed p.seqpercase
seed.df<-seed.df %>% mutate(seed.1 = ceiling((size2017 * 1/p.seqpercase * (1 - p.art) * (1 - p.suppressed) ) ))
seed.df
if(initialrun==FALSE){
  #use seed.adjust to adjust starting seed
  seed.adjust<-read.csv("2024-06-10_SBPresults/seed.adjust.csv")
  # seed.adjust join if not done already
  if(!any(colnames(seed.df)=="adjust.seed")){
    seed.df<-seed.df %>% left_join(seed.adjust,by="clusterID")
    seed.df$adjust.seed[which(is.na(seed.df$adjust.seed))]<-1
  }
  seed.df<-seed.df %>% mutate(seed.2=ceiling(seed.1*(1/adjust.seed)))
  ## Write the seed size
  write.csv(seed.df, paste0(f.out,"seed.df.csv"))
}
  
# max new cases (could tune to the cluster?)
max.samp<-500

#upper limits of infections and diagnoses rel to max samp
max.infect<-max.samp / p.caseperinfect / p.seqpercase
max.diag<-max.samp/ p.seqpercase
#consider delayed diagnoses in averted cases? Add 1 month (cite?)

#OUTPUT sims into a list (across clusters) of lists of dfs (one df per sim)
all.sims.out<-replicate(ncl, vector())
all.re.out<-replicate(ncl, vector()) #list of Re in sims
seedz<-c()

for (j in 1:ncl){
  set.seed(j)
  
  #prime the list of dfs
  run.df<-replicate(n=n.runs, vector())
  re.df<-replicate(n=n.runs,vector())

  # my cluster
  cl<-clust.unq[j]
  
  #initial seed of infectious cluster members
  #seed.1 is the inital guess. seed.2 is adjusted
  if(initialrun==TRUE) {seed<-seed.df$seed.1[seed.df$clusterID==cl]}
  if(initialrun==FALSE) {seed<-seed.df$seed.2[seed.df$clusterID==cl]}

  #minimum of 1 seed
  if(seed==0){seed<-1}
  seedz<-c(seedz,seed)
  
  # Re through time subset from above
  re.cl<-re.epoch %>% filter(clusterID==cl)
  
  for (i in 1:n.runs){
    set.seed(i+69) #seed for random number generator
    #initiate with seed form above
    cases<-seed
    t<-rep(0, seed)  
    times<-t #times/dates of infection
    res<-c() #re values over time
    reldayz<-c() #relative days when re looked up
    relday<-dayzero 
    
    #run until no more infected, more than max samp 500 (filter below), or at end date max long:
    while(cases > 0 & length(times)<max.infect & max(times)<max.long) { 
      
      #lookup epochal Re
      r.og<-r.now<-re.cl$Re[which(re.cl$date_start==(relday+1))]
      if(is.na(r.now)){r.now<-default.re} #default receding, shouldnt have to use this

      #for X% of generations, walk away from the epochal mean
      flip<-rbinom(1,1,0.6)

      ## according to gamma dist centered on epochal mean
      if(flip==1){
        r.z<-gammaParamsConvert(mean=r.now, sd=0.4)
        r.shape.now<-r.z$shape
        r.scale.now<-r.z$scale
        r.now<-rgamma(1, shape=r.shape.now, scale=r.scale.now)
      }
      #variability in dispersion
      k.now<-runif(n = 1, min = k.min, max=k.max)
      
      #variability in serial interval params
      si.shape.now<-runif(1, min = si.shape-0.01, max=si.shape+0.01)
      si.rate.now<-runif(1, min = si.rate-0.01, max=si.rate+0.01)
       
      # vector of the number of secondary cases infected by each infected case
      secondary<-rnbinom(cases, size= k.now, mu=r.now) 
      t.new<- numeric() 
      for(k in 1:length(secondary)) { 
        t.new<-c(t.new, t[k] + 
                   rgamma(secondary[k], shape=si.shape.now, rate=si.rate.now) )
      }
  
      #average time since origin of those infected. Use to lookup Re for next generation.
      relday<-round(mean(t.new))+dayzero # a date
      reldayz<-c(reldayz,relday)
      
      #number of actively infected cases
      cases<-length(t.new) 
      
      #t.new is a vector of infection times relative to beginning of simulation in this generation
      t<-t.new 
      times<-c(times,t.new) 
      res<-c(res,r.now) #only run if interested
    } #end of while loop
    
    #remove times that are greater than max (forward looking si)
    xs.times<-which(times>max.long)
    if(length(xs.times)>0){times<-times[-xs.times]}
    
    run.df[[i]]<-data.frame(order=1:length(times), infection.times=sort(times), run=i,
                            ass.rate=NA, diagnosed=0, seq.rate=NA, sampled=0) #default not diagnosed or sampled
    run.df[[i]]$clusterID<-cl
    
    re.df[[i]]<-data.frame(reldayz, Re=res)
    # re.df$[[i]]$clusterID<-cl
    
    #draw an ascertainment rate (diagnosed cases per infected)
    ass.rate<-rlnorm(1,mean=log(p.caseperinfect), sd=0.02)
    if(ass.rate>0.99) {ass.rate<-0.99} #set max #needed if applying weights or wide sd
    if(ass.rate<0.5) {ass.rate<-0.5} #set min
  
    #who was diagnosed in DTP?
    run.df[[i]]$ass.rate<-ass.rate #log it
    sz.inf<-nrow(run.df[[i]])  
    run.df[[i]]$diagnosed<-rbinom(n=sz.inf, size=1, prob = ass.rate)
    
    #Reduce memory by only including those diagnosed (total infected irrelevant for analyses)
    run.df[[i]]<-run.df[[i]] %>% filter(diagnosed==1)
    
    #who was sampled/sequenced given diagnosed?
    sz.diag<-nrow(run.df[[i]])
    
    if(sz.diag>0){
      seq.rate<-rlnorm(1,mean=log(p.seqpercase), sd=0.1)
      if(seq.rate>0.99){seq.rate<-0.99} #impose a max
      run.df[[i]]$seq.rate<-seq.rate
      run.df[[i]]$sampled<-rbinom(n=sz.diag, size=1, prob = seq.rate)
    }
  } #end of i loop of n runs for each cluster
  
  # export lists of dfs as elements of lists
  all.sims.out[[j]]<-run.df
  all.re.out[[j]]<-re.df
  
} #end of j loop across clust


# For observed Re, back up
if(type.re=="Re_observed"){all.sims.out.obs<-all.sims.out}
```

## Filter simulations
```{r}
#iterate through clusters to filter empty, too big, ...
filter.simlist<-function(list){
  for(j in 1:ncl){
    ## remove empty elements inthe list (if memory limit exhausted, or stop early)
    list[[j]]<-list_drop_empty(list[[j]]) #this also removes any with no diagnoses only infected
    length(list[[j]])
    all.re.out[[j]]<-list_drop_empty(all.re.out[[j]])
    length(all.re.out[[j]])
    
    #Id and remove any simulations with these criteria
    zers<-c()
    highsamp<-c()
    highinf<-c()
    
    #iterate across each simulation 
    for (i in 1:length(list[[j]])){ 
      #how many were sampled total in this sim
      ll<-length(which(list[[j]][[i]]$sampled==1))
      #with zero sampled
      if(ll<1){
        zers<-c(zers, i)
      }
      #no bigger n.sample than 1.1x biggest sublineage that was observed
      if(ll>(max.samp*1.1)){
        highsamp<-c(highsamp,i)
      }
      #total infected greater than max total.size (outbreak stopped short here)
      ln<-nrow(list[[j]][[i]]) #nrow is one for every infected
      if(ln>max.infect){
        highinf<-c(highinf,i)
      }
    }
    
    length(zers) 
    length(highsamp)
    length(highinf)
    
    goners<-unique(c(zers,highsamp,highinf)) # 
    length(goners)
    
    #remove i goners/sims
    if(length(goners)>0){
      list[[j]]<-list[[j]][-goners]
    }
    
  } #end of loop  
  return(list)
}

# all.sims.out.obs<-all.sims.out
all.sims.out<-filter.simlist(all.sims.out)
all.sims.out.obs<-filter.simlist(all.sims.out.obs)
```

## bind simulations by cluster ##
```{r}
run.df.all<-replicate(ncl, vector())
run.df.summ<-replicate(ncl, vector())

for (j in 1:ncl){
  #run.df.all is every case in every simulation
  run.df.all[[j]]<-dplyr::bind_rows(all.sims.out[[j]])
  
  #run.df.summ is one row per simulation run
  run.df.summ[[j]]<-run.df.all[[j]] %>% dplyr::group_by(run) %>%
    dplyr::summarize(clusterID=clusterID[1],
                     n=n(),
                     n.diag=length(which(diagnosed==1)),
                     n.sample=length(which(sampled==1)),
                     max.time=max(infection.times))
}

#observed sims
run.df.all.obs<-replicate(ncl, vector())
run.df.summ.obs<-replicate(ncl, vector())

for (j in 1:ncl){
  #run.df.all is every case in every simulation
  run.df.all.obs[[j]]<-dplyr::bind_rows(all.sims.out.obs[[j]])
  
  #run.df.summ is one row per simulation run
  run.df.summ.obs[[j]]<-run.df.all.obs[[j]] %>% dplyr::group_by(run) %>%
    dplyr::summarize(clusterID=clusterID[1],
                     n=n(),
                     n.diag=length(which(diagnosed==1)),
                     n.sample=length(which(sampled==1)),
                     max.time=max(infection.times))
}

#no PrEP fit
run.df.summ.big<-bind_rows(run.df.summ)
#observed/PrEP fit
run.df.summ.big.obs<-bind_rows(run.df.summ.obs)
```


## Make a plot of N diag predicted and observed, facetted by cluster
```{r}
# Add mean n case
run.df.summ.big.summ<-run.df.summ.big %>% 
  # filter(clusterID %in% med.active.clust)%>%
  group_by(clusterID)%>%
  dplyr::summarize(mean.diag=round(mean(n.diag, na.rm=T)),
                   #quantiles for CI isn't right...
                   upper.diag=round(quantile(n.diag,0.95, na.rm=T)),
                   lower.diag=round(quantile(n.diag,0.05, na.rm=T)) )

# Add bootstrap CIs (resample 95% of data, calculate and track mean, repeat X times, then calc regular CI)
n.boot<-999
getmean <- function(data, indices) {
  d <- data[indices,] 
  return(mean(d$n.diag,na.rm=T))
}
df.boot<-data.frame(clusterID=clust.unq, mean=NA, upper=NA, lower=NA)
for (i in 1:ncl){
  df<-run.df.summ.big %>% filter(clusterID == clust.unq[i])
    boot.1<-boot(df, statistic=getmean, R = 1000)
  b<-boot.ci(boot.1,conf = 0.95,type = "norm") %>% unlist()
  df.boot$mean[df.boot$clusterID == clust.unq[i]]<-b$t0 %>% round(digits = 1)
  df.boot$upper[df.boot$clusterID == clust.unq[i]]<-b$normal3 %>% round(digits = 1)
  df.boot$lower[df.boot$clusterID == clust.unq[i]]<-b$normal2 %>% round(digits = 1)
}

#join on to the run.df.summ.big.summ
run.df.summ.big.summ<-run.df.summ.big.summ %>% left_join(df.boot, by="clusterID")
head(run.df.summ.big.summ)

#text that combines mean and confints
run.df.summ.big.summ<-run.df.summ.big.summ %>% 
  unite(col=ci, lower, upper, sep = "-", remove = F) %>%
  unite(col=text, mean, ci, sep = " (", remove = F)
run.df.summ.big.summ$text<-paste0(run.df.summ.big.summ$text,")")

run.df.summ.big.summ.medact<-run.df.summ.big.summ %>% 
  filter(clusterID %in% med.active.clust) 
# run.df.summ.big.summ.medact.obs<-run.df.summ.big.summ.obs %>% 
#   filter(clusterID %in% med.active.clust) 

if(type.re=="Re_observed"){
  plotcolor<-prepcolz[1]
  textlabel<-"PrEP (sim.): "
}
if(type.re!="Re_observed"){
  plotcolor<-prepcolz[2]
  textlabel<-"no PrEP (sim.): "
}

# Plot of predicted distrib with observed line, both annotated
#PLOT
P3.bar.sim.diag.f<-run.df.summ.big %>% 
  filter(clusterID %in% med.active.clust) %>%
    ggplot()+
     
    geom_bar(aes(x=n.diag),stat="count", fill=plotcolor, alpha=0.8)+
    geom_vline(data=run.df.summ.big.summ.medact, 
               aes(xintercept = mean.diag, group=clusterID), color=plotcolor,linetype=6)+
    geom_text(data=run.df.summ.big.summ.medact, aes(x = mean.diag, y=1,group=clusterID,
                                             label=paste0(textlabel, text)),
              color=plotcolor,size=2.4,vjust=-3,hjust=-0.05)+

    #add a line for the observed value in different colour
    geom_vline(data=clust.obs.cum.medact, aes(xintercept = total.observed, group=clusterID), color = prepcolz[1],linetype=6)+
    geom_text(data=clust.obs.cum.medact, aes(x = total.observed,  y=1,group=clusterID, 
                                             label=paste0("PrEP (obs.): ",ceiling(total.observed))),
              color=prepcolz[1],size=2.4,vjust=-1,hjust=-0.05)+
  
    pubTheme+
    scale_x_continuous(expand=expansion(mult=c(0,1)))+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          axis.text.x=element_text(angle=45,vjust=1,hjust=1,size=rel(1.2)),
          # legend.position = c(0.7,0.8),
          legend.position = "none",
          legend.text = element_text(size=9),
          # axis.title.y=element_text(vjust=4),       
          axis.title.y=element_text(vjust=1,size=rel(1.2)), 
          plot.margin=unit(c(4,4,4,4),"pt"))+
    labs(x="n diagnosed",y="Count")+
    guides(fill=guide_legend(ncol= 1,title="Lineage",title.position="top",reverse = T, keywidth = 1.2,keyheight = 1.2))+
  facet_wrap(~clusterID, scales="free",nrow=6)

P3.bar.sim.diag.f
ggsave(paste0(f.out, type.re, "_simulations_Ndiag.png"), width = 8.5,height=7)
```

## Repeat for all clustres, not just active med
```{r}

# Plot of predicted distrib with observed line, both annotated
#PLOT
P3.bar.sim.diag.f.mo<-run.df.summ.big %>% 
    ggplot()+
    geom_bar(aes(x=n.diag),stat="count", fill=plotcolor, alpha=0.8)+
    geom_vline(data=run.df.summ.big.summ, 
               aes(xintercept = mean.diag, group=clusterID), color=plotcolor,linetype=6)+
    geom_text(data=run.df.summ.big.summ, aes(x = mean.diag, y=1,group=clusterID,
                                             label=paste0(textlabel, text)),
              color=plotcolor,size=2.4,vjust=-3,hjust=-0.05)+

    #add a line for the observed value in different colour
    geom_vline(data=clust.obs.cum, aes(xintercept = total.observed, group=clusterID), color = prepcolz[1],linetype=6)+
    geom_text(data=clust.obs.cum, aes(x = total.observed,  y=1,group=clusterID, 
                                             label=paste0("PrEP (obs.): ",ceiling(total.observed))),
              color=prepcolz[1],size=2.4,vjust=-1,hjust=-0.05)+
  
    pubTheme+
    scale_x_continuous(expand=expansion(mult=c(0,1)))+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          axis.text.x=element_text(angle=45,vjust=1,hjust=1,size=rel(1.2)),
          # legend.position = c(0.7,0.8),
          legend.position = "none",
          legend.text = element_text(size=9),
          # axis.title.y=element_text(vjust=4),       
          axis.title.y=element_text(vjust=1,size=rel(1.2)), 
          plot.margin=unit(c(4,4,4,4),"pt"))+
    labs(x="n diagnosed",y="Count")+
    guides(fill=guide_legend(ncol= 1,title="Lineage",title.position="top",reverse = T, keywidth = 1.2,keyheight = 1.2))+
  facet_wrap(~clusterID, scales="free")

P3.bar.sim.diag.f.mo
ggsave(paste0(f.out, type.re, "_simulations_Ndiag_mo.png"), width = 12,height=15)
```

# Repeat for samples (apples to apples b/c cluster members are diagnosed and sequenced)
```{r}
# Add median n case
run.df.summ.big.summ.samp<-run.df.summ.big %>% 
  # filter(clusterID %in% med.active.clust)%>%
  group_by(clusterID)%>%
  dplyr::summarize(mean.sample=round(mean(n.sample, na.rm=T)),
                   upper.sample=round(quantile(n.sample,0.95, na.rm=T)),
                   lower.sample=round(quantile(n.sample,0.05, na.rm=T)) )

# Add bootstrap CIs (resample 95% of data, calculate and track mean, repeat X times, then calc regular CI)
n.boot<-999
getmean.samp <- function(data, indices) {
  d <- data[indices,] 
  return(mean(d$n.sample,na.rm=T)) }
df.boot<-data.frame(clusterID=clust.unq, mean=NA, upper=NA, lower=NA)
for (i in 1:ncl){
  df<-run.df.summ.big%>% filter(clusterID == clust.unq[i])
    boot.1<-boot(df, statistic=getmean.samp, R = 1000)
  b<-boot.ci(boot.1,conf = 0.95,type = "norm") %>% unlist()
  df.boot$mean[df.boot$clusterID == clust.unq[i]]<-b$t0 %>% round(digits = 1)
  df.boot$upper[df.boot$clusterID == clust.unq[i]]<-b$normal3 %>% round(digits = 1)
  df.boot$lower[df.boot$clusterID == clust.unq[i]]<-b$normal2 %>% round(digits = 1)
}

#join on to the run.df.summ.big.summ
run.df.summ.big.summ.samp<-run.df.summ.big.summ.samp %>% left_join(df.boot, by="clusterID")

#text that combines mean and confints
run.df.summ.big.summ.samp<-run.df.summ.big.summ.samp %>% 
  unite(col=ci, lower, upper, sep = "-", remove = F) %>%
  unite(col=text, mean, ci, sep = " (", remove = F)
run.df.summ.big.summ.samp$text<-paste0(run.df.summ.big.summ.samp$text,")")

# Plot of predicted distrib with observed line, both annotated
run.df.summ.big.summ.samp.medact<-run.df.summ.big.summ.samp %>%
  filter(clusterID %in% med.active.clust)
# run.df.summ.big.summ.samp.medact.obs<-run.df.summ.big.summ.samp.obs %>%
#   filter(clusterID %in% med.active.clust)

#PLOT
P3.bar.sim.sample.f.samp<-run.df.summ.big %>% 
  filter(clusterID %in% med.active.clust) %>%
    ggplot()+
    geom_bar(aes(x=n.sample),stat="count", fill=plotcolor, alpha=0.8)+
    geom_vline(data=run.df.summ.big.summ.samp.medact, 
               aes(xintercept = mean.sample, group=clusterID), color=plotcolor,linetype=6)+
    geom_text(data=run.df.summ.big.summ.samp.medact, 
              aes(x = mean.sample, y=1,group=clusterID,
                                             label=paste0(textlabel,text)),
              color=plotcolor,size=2.4,vjust=-3,hjust=-0.05)+
 
    #add a line for the observed value in different colour
    geom_vline(data=clust.obs.cum.medact, aes(xintercept = total.observed, group=clusterID), color = prepcolz[1],linetype=6)+
    geom_text(data=clust.obs.cum.medact, aes(x = total.observed,  y=1,group=clusterID, 
                                             label=paste0("PrEP (obs.): ",ceiling(total.observed))),
              color=prepcolz[1],size=2.4,vjust=-1,hjust=-0.05)+

    pubTheme+
    scale_x_continuous(expand=expansion(mult=c(0,1)))+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          axis.text.x=element_text(angle=45,vjust=1,hjust=1,size=rel(1.2)),
          # legend.position = c(0.7,0.8),
          legend.position = "none",
          legend.text = element_text(size=9),
          # axis.title.y=element_text(vjust=4),       
          axis.title.y=element_text(vjust=1,size=rel(1.2)), 
          plot.margin=unit(c(4,4,4,4),"pt"))+
    labs(x="n sampled",y="Count")+
    # guides(fill=guide_legend(ncol= 1,title="Lineage",title.position="top",reverse = T, keywidth = 1.2,keyheight = 1.2))+
  facet_wrap(~clusterID, scales="free",nrow=6)

P3.bar.sim.sample.f.samp
ggsave(paste0(f.out, type.re, "_simulations_Nsample.png"), width = 8.5,height=7)
```

## Seed adjustment
## calculate diff between expected n sampled (using Re observed) and observed n case
```{r}
if(type.re=="Re_observed" & initialrun==TRUE){
  seed.adjust<-left_join(clust.obs.cum, run.df.summ.big.summ.samp, by="clusterID")
  seed.adjust<-seed.adjust %>% dplyr::select(clusterID, total.observed, mean.sample)
  seed.adjust<-seed.adjust %>% mutate(adjust.seed=mean.sample/total.observed)
  # if >1 then too big in sim; if <1, too small #mult by 1/adjust.seed
  
  #could also adjust Re by this value... but I think more likely related to misspecified seed
  write.csv(seed.adjust,paste0(f.out, "seed.adjust.csv"))
}

```


## Join simulated PrEP and no PREP to calculate cases (samples and diagnoses) averted
```{r}
colnames(clust.obs.cum)
run.df.summ.big.obs.r<-run.df.summ.big.obs[,-c(3,6)]
run.df.summ.big.r<-run.df.summ.big[,-c(3,6)]

#add type 
clust.obs.cum$type<-"PrEP_obs"
run.df.summ.big.obs.r$type<-"PrEP_sim"
run.df.summ.big.r$type<-"NoPrEP_sim"

#bind rows
run.df.summ.big.both.long<-bind_rows(run.df.summ.big.r, run.df.summ.big.obs.r)


# #join by cluster ID (and run?)
# run.df.summ.big.both<-left_join(run.df.summ.big.r, run.df.summ.big.obs.r, by=c("clusterID","run"))
# colnames(run.df.summ.big.both)

# #make long
# run.df.summ.big.both.long<-run.df.summ.big.both %>%
#   pivot_longer(cols = )

temp<-run.df.summ.big.both.long %>%
  # filter(clusterID %in% med.active.clust) %>%
  dplyr::group_by(clusterID, type) %>%
  dplyr::summarise(clusterID=clusterID[1],
                   type=type[1],
                   mean=mean(log(n.diag)))

#SORT these in order of most averted like the other plot
run.df.summ.big.both.long$clusterID<-factor(run.df.summ.big.both.long$clusterID, levels=clust.ord2)
temp$clusterID<-factor(temp$clusterID, levels = clust.ord2)
run.df.summ.big.both.long$type2<-str_replace_all(run.df.summ.big.both.long$type, c("NoPrEP_sim"="No PrEP",
                                                                                   "PrEP_sim"="PrEP since 2018"))
temp$type2<-str_replace_all(temp$type, c("NoPrEP_sim"="No PrEP","PrEP_sim"="PrEP since 2018"))
  
#Plot the distrib diagnoses in either by cluster
p.sims<-run.df.summ.big.both.long %>%
  # filter(clusterID %in% med.active.clust) %>%
  ggplot()+
  geom_boxplot(aes(x=as.factor(clusterID), y=log(n.diag), color=type2),
             position = position_dodge(width=0.8),
             width=0.5,alpha=0.4,outlier.alpha = 0.1, fatten=NULL,size=0.4,outlier.size=0.4)+
  geom_point(data=temp, aes(x=as.factor(clusterID), y=mean, color=type2),position = position_dodge(width=0.8),shape="-",size=4)+
  pubTheme+
  theme(legend.position = c(0.35,1), legend.justification = c(0.5,1),
        axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  scale_color_manual(values=rev(as.character(prepcolz)))+
  guides(color=guide_legend(title="Simulation scenario"))+
  labs(x="Cluster", y="log N diagnosed")+
  scale_y_continuous(expand=c(0.1,0), limits = c(0,8))
p.sims
ggsave(paste0(f.out,"Simulated_PrEP-NoPrEP_Log-NDiag.png"),width=8.5,height=5)
# 
# 
# #Now samples
# run.df.summ.big.both.long %>%
#   filter(clusterID %in% med.active.clust) %>%
#   ggplot()+
#   geom_boxplot(aes(x=as.factor(clusterID), y=n.sample, color=type),
#              position = position_dodge(width=0.8),
#              width=0.5,alpha=0.4)+
#   pubTheme+
#   theme(legend.position = c(0.9,0.95), legend.justification = c(1,1))+
#   scale_color_manual(values=rev(as.character(prepcolz)))+
#   labs(x="Cluster", y="N sampled")
# ggsave(paste0(f.out,"Simulated_PrEP-NoPrEP_NSamp.png"),width=8.5,height=4.5)

```

## Calculate samples and diagnoses averted 
```{r}
## Create summary objects for the obs data
run.df.summ.big.summ.obs<-run.df.summ.big.obs %>% 
  group_by(clusterID)%>%
  dplyr::summarize(mean.diag=round(mean(n.diag, na.rm=T)),
                   upper.diag=round(quantile(n.diag,0.95, na.rm=T)),
                   lower.diag=round(quantile(n.diag,0.05, na.rm=T)) )

# Add bootstrap CIs (resample 95% of data, calculate and track mean, repeat X times, then calc regular CI)
df.boot<-data.frame(clusterID=clust.unq, mean=NA, upper=NA, lower=NA)
for (i in 1:ncl){
  df<-run.df.summ.big.obs %>% filter(clusterID == clust.unq[i])
    boot.1<-boot(df, statistic=getmean, R = 1000)
  b<-boot.ci(boot.1,conf = 0.95,type = "norm") %>% unlist()
  df.boot$mean[df.boot$clusterID == clust.unq[i]]<-b$t0 %>% round(digits = 1)
  df.boot$upper[df.boot$clusterID == clust.unq[i]]<-b$normal3 %>% round(digits = 1)
  df.boot$lower[df.boot$clusterID == clust.unq[i]]<-b$normal2 %>% round(digits = 1)
}

#join on to the run.df.summ.big.summ
run.df.summ.big.summ.obs<-run.df.summ.big.summ.obs %>% left_join(df.boot, by="clusterID")

#text that combines mean and confints
run.df.summ.big.summ.obs<-run.df.summ.big.summ.obs %>% 
  unite(col=ci, lower, upper, sep = "-", remove = F) %>%
  unite(col=text, mean, ci, sep = " (", remove = F)
run.df.summ.big.summ.obs$text<-paste0(run.df.summ.big.summ.obs$text,")")
# head(run.df.summ.big.summ.obs)

# Add mean n case
run.df.summ.big.summ.samp.obs<-run.df.summ.big.obs %>% 
  group_by(clusterID)%>%
  dplyr::summarize(mean.sample=round(mean(n.sample, na.rm=T)),
                   upper.sample=round(quantile(n.sample,0.95, na.rm=T)),
                   lower.sample=round(quantile(n.sample,0.05, na.rm=T)) )

# Add bootstrap CIs (resample 95% of data, calculate and track mean, repeat X times, then calc regular CI)
df.boot<-data.frame(clusterID=clust.unq, mean=NA, upper=NA, lower=NA)
for (i in 1:ncl){
  df<-run.df.summ.big.obs%>% filter(clusterID == clust.unq[i])
    boot.1<-boot(df, statistic=getmean.samp, R = 1000)
  b<-boot.ci(boot.1,conf = 0.95,type = "norm") %>% unlist()
  df.boot$mean[df.boot$clusterID == clust.unq[i]]<-b$t0 %>% round(digits = 1)
  df.boot$upper[df.boot$clusterID == clust.unq[i]]<-b$normal3 %>% round(digits = 1)
  df.boot$lower[df.boot$clusterID == clust.unq[i]]<-b$normal2 %>% round(digits = 1)
}

#join on to the run.df.summ.big.summ
run.df.summ.big.summ.samp.obs<-run.df.summ.big.summ.samp.obs %>% left_join(df.boot, by="clusterID")

#text that combines mean and confints
run.df.summ.big.summ.samp.obs<-run.df.summ.big.summ.samp.obs %>% 
  unite(col=ci, lower, upper, sep = "-", remove = F) %>%
  unite(col=text, mean, ci, sep = " (", remove = F)
run.df.summ.big.summ.samp.obs$text<-paste0(run.df.summ.big.summ.samp.obs$text,")")


## COMBINE
samp.prepsim.summ<-run.df.summ.big.summ.samp.obs
samp.noprepsim.summ<-run.df.summ.big.summ.samp

colnames(samp.prepsim.summ)[-1]<-paste0(colnames(samp.prepsim.summ)[-1], "_prepsim")
colnames(samp.noprepsim.summ)[-1]<-paste0(colnames(samp.noprepsim.summ)[-1], "_noprepsim")

#JOIN EM
run.df.summ.big.both.samp<-left_join(samp.prepsim.summ, samp.noprepsim.summ, by=c("clusterID"))

# Calculate total samples averted (relative to the simulated data fitted to observed)
run.df.summ.big.both.samp<-run.df.summ.big.both.samp %>%
  mutate(mean.sample.averted =  mean_noprepsim - mean_prepsim, 
         lower.sample.averted = lower_noprepsim - lower_prepsim  ,
         upper.sample.averted = upper_noprepsim- upper_prepsim  )
# run.df.summ.big.both.samp$mean.sample.averted

#Join cluster comp on 
run.df.summ.big.both.samp.2<-run.df.summ.big.both.samp %>% left_join (clust.comp, by="clusterID")

clust.ord1 <- run.df.summ.big.both.samp.2$clusterID [order(run.df.summ.big.both.samp.2$mean.sample.averted,decreasing = T)]
run.df.summ.big.both.samp.2$clusterID<-factor(run.df.summ.big.both.samp.2$clusterID, levels=clust.ord1)
#Make plot with ranges for each
yrange<-c(-20,30)

##PLOT THIS
run.df.summ.big.both.samp.2%>%
  filter(clusterID %in% med.active.clust) %>%
  ggplot()+
  geom_bar(aes(x=as.factor(clusterID), y=mean.sample.averted, fill=risk), stat="identity",alpha=0.7)+
  geom_errorbar(aes(x=as.factor(clusterID), ymin=lower.sample.averted, ymax=upper.sample.averted, color=risk), 
                alpha=0.7, width = 0.3)+
  pubTheme+
  scale_fill_manual(values=riskcolz)+
  scale_color_manual(values=riskcolz,guide="none")+
  labs(x="Active clusters",y="Samples averted (sim. no PreP - sim. with PrEP)")+
  guides(fill=guide_legend(title="Predominant population"))+
  theme(legend.position = c(0.95,0.95), legend.justification = c(1,1),
        axis.text.x=element_text(angle=45,hjust=1,vjust=1)
        )+
  scale_y_continuous(breaks=seq(yrange[1],yrange[2],10),expand=c(0.1,0))+
  coord_cartesian(ylim=yrange)
ggsave(paste0(f.out,"/SamplesAvertedByRiskGrp.png"),width=8,height=4)

## Repeat for diagnoses
diag.prepsim.summ<-run.df.summ.big.summ.obs
diag.noprepsim.summ<-run.df.summ.big.summ

colnames(diag.prepsim.summ)[-1]<-paste0(colnames(diag.prepsim.summ)[-1], "_prepsim")
colnames(diag.noprepsim.summ)[-1]<-paste0(colnames(diag.noprepsim.summ)[-1], "_noprepsim")

#JOIN EM
run.df.summ.big.both.diag<-left_join(diag.prepsim.summ, diag.noprepsim.summ, by=c("clusterID"))

# Calculate total Diagnoses averted (relative to the simulated data fitted to observed)
run.df.summ.big.both.diag<-run.df.summ.big.both.diag %>%
  mutate(mean.diag.averted = mean_noprepsim- mean_prepsim, 
         lower.diag.averted =lower_noprepsim - lower_prepsim  ,
         upper.diag.averted = upper_noprepsim-upper_prepsim)
run.df.summ.big.both.diag$mean.diag.averted

#Join cluster comp on 
run.df.summ.big.both.diag.2<-run.df.summ.big.both.diag %>% left_join (clust.comp, by="clusterID")

#Make plot with ranges for each
yrange<-c(-21,30)

#new cluster order highest to lowest
clust.ord2 <- run.df.summ.big.both.diag.2$clusterID [order(run.df.summ.big.both.diag.2$mean.diag.averted,decreasing = T)]
run.df.summ.big.both.diag.2$clusterID<-factor(run.df.summ.big.both.diag.2$clusterID, levels=clust.ord2)

##PLOT THIS
p.averted<-run.df.summ.big.both.diag.2%>%
  # filter(clusterID %in% med.active.clust) %>%
  ggplot()+
  geom_bar(aes(x=as.factor(clusterID), y=mean.diag.averted, fill=risk), stat="identity",alpha=0.7)+
  geom_errorbar(aes(x=as.factor(clusterID), ymin=lower.diag.averted, ymax=upper.diag.averted, color=risk), 
                alpha=0.7, width = 0.3)+
  pubTheme+
  scale_fill_manual(values=riskcolz)+
  scale_color_manual(values=riskcolz,guide="none")+
  labs(x="Cluster",y="Diagnoses averted by PrEP")+
  guides(fill=guide_legend(title="Predominant population"))+
  theme(legend.position = c(0.35,1), legend.justification = c(0.5,1),
        axis.text.x=element_text(angle=45,hjust=1,vjust=1)
        )+
  scale_y_continuous(breaks=seq(yrange[1],yrange[2],10),expand=c(0.01,0))+
  coord_cartesian(ylim=yrange)
p.averted

ggsave(paste0(f.out,"/DiagnosesAvertedByRiskGrp.png"),width=8.5,height=4)

```

## grob this together with diag averted across sims
```{r}
p.sims.nox<-p.sims+theme(axis.title.x=element_blank())
pz<-plot_grid(p.sims.nox,p.averted, labels=c("A","B"), align="v",nrow=2)
pz
ggsave(paste0(f.out, "SimulationsScenarios_DiagAverted.png"),width=9, height=6)
```

## cor?
```{r}
p.averted.age<-run.df.summ.big.both.diag.2%>%
  filter(clusterID %in% med.active.clust) %>%
  ggplot()+
  geom_point(aes(x=as.factor(clusterID), y=mean.diag.averted, fill=risk), stat="identity",alpha=0.7)+
  
  pubTheme+
  scale_fill_manual(values=riskcolz)+
  scale_color_manual(values=riskcolz,guide="none")+
  labs(x="Cluster",y="Diagnoses averted by PrEP")+
  guides(fill=guide_legend(title="Predominant population"))+
  theme(legend.position = c(0.4,0.999), legend.justification = c(0.5,1),
        axis.text.x=element_text(angle=45,hjust=1,vjust=1)
        )+
  scale_y_continuous(breaks=seq(yrange[1],yrange[2],10),expand=c(0.1,0))+
  coord_cartesian(ylim=yrange)
p.averted
```


# data queries
```{r}
#clusters with most averted diagnoses
run.df.summ.big.both.diag.2$clusterID[run.df.summ.big.both.diag.2$mean.diag.averted>0]
run.df.summ.big.both.diag.2$clusterID[order(run.df.summ.big.both.diag.2$mean.diag.averted, decreasing = T)]
run.df.summ.big.both.diag.2$clusterID[which(run.df.summ.big.both.diag.2$mean.diag.averted>=1)]

# If we assume no negative cases averted...how many total averted diagnoses
total.diag.averted<-run.df.summ.big.both.diag.2 %>%
  filter(mean.diag.averted>0)%>%
  dplyr::summarise(sumdiagavert=sum(mean.diag.averted),
                   sumdiagavert.low=sum(lower.diag.averted),
                   sumdiagavert.up=sum(upper.diag.averted))
total.diag.averted$sumdiagavert #107.6
total.diag.averted$sumdiagavert.low #99.1
total.diag.averted$sumdiagavert.up #115.7

#diagnoses averted per year from 2018 to 2022
total.diag.averted$sumdiagavert/max.long*365 #21.52
total.diag.averted$sumdiagavert.low/max.long*365 #19.8
total.diag.averted$sumdiagavert.up/max.long*365 #23.1

# If we assume no negative cases averted...how many total averted samples
total.samp.averted<-run.df.summ.big.both.samp.2 %>%
  filter(mean.sample.averted>0)%>%
  dplyr::summarise(sumsampavert=sum(mean.sample.averted),
                   sumsampavert.low=sum(lower.sample.averted),
                   sumsampavert.up=sum(upper.sample.averted))
total.samp.averted$sumsampavert #98.2
total.samp.averted$sumsampavert.low #91.4
total.samp.averted$sumsampavert.up #106.7

#diagnoses averted per year from 2018 to 2022
total.samp.averted$sumsampavert/max.long*365 #19.8
total.samp.averted$sumsampavert.low/max.long*365 #18.3
total.samp.averted$sumsampavert.up/max.long*365 #21.3

## Key clusters
#CLUSTER 22
run.df.summ.big.both.diag.2$medianAge2023 [run.df.summ.big.both.diag.2$clusterID==22] # 44.3
run.df.summ.big.both.diag.2$medianAgeFARV [run.df.summ.big.both.diag.2$clusterID==22] # 39
run.df.summ.big.both.diag.2$perc_MSM [run.df.summ.big.both.diag.2$clusterID==22] # 100
run.df.summ.big.both.diag.2$perc_HA_Van [run.df.summ.big.both.diag.2$clusterID==22] # 73.6
run.df.summ.big.both.diag.2$perc_PREP [run.df.summ.big.both.diag.2$clusterID==22] # 26.3
run.df.summ.big.both.diag.2$n_seroconv_2018 [run.df.summ.big.both.diag.2$clusterID==22] # 19
run.df.summ.big.both.diag.2$n_PREP [run.df.summ.big.both.diag.2$clusterID==22] # 5
run.df.summ.big.both.diag.2$text_noprepsim [run.df.summ.big.both.diag.2$clusterID==22] # "49.9 (45.9-54.1)"
run.df.summ.big.both.diag.2$text_prepsim [run.df.summ.big.both.diag.2$clusterID==22] # "24 (21.7-26.4)"
run.df.summ.big.both.diag.2$mean.diag.averted [run.df.summ.big.both.diag.2$clusterID==22] # "25.9
run.df.summ.big.both.diag.2$lower.diag.averted [run.df.summ.big.both.diag.2$clusterID==22] # "24.2
run.df.summ.big.both.diag.2$upper.diag.averted [run.df.summ.big.both.diag.2$clusterID==22] # "27.7

#CLUSTER 31
run.df.summ.big.both.diag.2$medianAge2023 [run.df.summ.big.both.diag.2$clusterID==31] # 43.6
run.df.summ.big.both.diag.2$medianAgeFARV [run.df.summ.big.both.diag.2$clusterID==31] # 35
run.df.summ.big.both.diag.2$perc_MSM [run.df.summ.big.both.diag.2$clusterID==31] # 88.9
run.df.summ.big.both.diag.2$perc_HA_Van [run.df.summ.big.both.diag.2$clusterID==31] # 60.5
run.df.summ.big.both.diag.2$perc_PREP [run.df.summ.big.both.diag.2$clusterID==31] # 8.5
run.df.summ.big.both.diag.2$n_seroconv_2018 [run.df.summ.big.both.diag.2$clusterID==31] # 94
run.df.summ.big.both.diag.2$n_PREP [run.df.summ.big.both.diag.2$clusterID==31] #8
run.df.summ.big.both.diag.2$text_noprepsim [run.df.summ.big.both.diag.2$clusterID==31] # "115.7 (112.3-119.1)"
run.df.summ.big.both.diag.2$text_prepsim [run.df.summ.big.both.diag.2$clusterID==31] #  "86.8 (84.2-89.4)"
run.df.summ.big.both.diag.2$mean.diag.averted [run.df.summ.big.both.diag.2$clusterID==31] # "28.9
run.df.summ.big.both.diag.2$lower.diag.averted [run.df.summ.big.both.diag.2$clusterID==31] # "28.1
run.df.summ.big.both.diag.2$upper.diag.averted [run.df.summ.big.both.diag.2$clusterID==31] # "29.7

# 209 (PWID)
run.df.summ.big.both.diag.2$text_noprepsim [run.df.summ.big.both.diag.2$clusterID==209] #  "13.4 (12-14.7)"
run.df.summ.big.both.diag.2$text_prepsim [run.df.summ.big.both.diag.2$clusterID==209] # "24 (21.7-26.4)"
run.df.summ.big.both.diag.2$mean.diag.averted [run.df.summ.big.both.diag.2$clusterID==209] # "25.9
run.df.summ.big.both.diag.2$lower.diag.averted [run.df.summ.big.both.diag.2$clusterID==209] # "24.2
run.df.summ.big.both.diag.2$upper.diag.averted [run.df.summ.big.both.diag.2$clusterID==209] # "27.7
#31
run.df.summ.big.both.diag.2$text_noprepsim [run.df.summ.big.both.diag.2$clusterID==31] # "115.7 (112.3-119.1)"
run.df.summ.big.both.diag.2$text_prepsim [run.df.summ.big.both.diag.2$clusterID==31] #  "86.8 (84.2-89.4)"
run.df.summ.big.both.diag.2$mean.diag.averted [run.df.summ.big.both.diag.2$clusterID==31] # "28.9
run.df.summ.big.both.diag.2$lower.diag.averted [run.df.summ.big.both.diag.2$clusterID==31] # "28.1
run.df.summ.big.both.diag.2$upper.diag.averted [run.df.summ.big.both.diag.2$clusterID==31] # "29.7

#what are charactersitcs of negative cases averted 
# 142, 234, and 13. 
run.df.summ.big.both.diag.2$medianAge2023 [run.df.summ.big.both.diag.2$clusterID==142] #55
run.df.summ.big.both.diag.2$medianAgeFARV [run.df.summ.big.both.diag.2$clusterID==142] #43
run.df.summ.big.both.diag.2$perc_MSM [run.df.summ.big.both.diag.2$clusterID==142] # 75
run.df.summ.big.both.diag.2$perc_HA_Van [run.df.summ.big.both.diag.2$clusterID==142] # 32.7
run.df.summ.big.both.diag.2$perc_HA_VanIsl [run.df.summ.big.both.diag.2$clusterID==142] # 56.4
run.df.summ.big.both.diag.2$perc_PREP [run.df.summ.big.both.diag.2$clusterID==142] # 0
run.df.summ.big.both.diag.2$n_seroconv_2018 [run.df.summ.big.both.diag.2$clusterID==142] # 6

run.df.summ.big.both.diag.2$medianAge2023 [run.df.summ.big.both.diag.2$clusterID==234] #44.3
run.df.summ.big.both.diag.2$medianAgeFARV [run.df.summ.big.both.diag.2$clusterID==234] #33
run.df.summ.big.both.diag.2$perc_MSM [run.df.summ.big.both.diag.2$clusterID==234] # 78.8
run.df.summ.big.both.diag.2$perc_HA_Van [run.df.summ.big.both.diag.2$clusterID==234] # 30.8
run.df.summ.big.both.diag.2$perc_HA_VanIsl [run.df.summ.big.both.diag.2$clusterID==234] # 12.8
run.df.summ.big.both.diag.2$perc_PREP [run.df.summ.big.both.diag.2$clusterID==234] # 16.7
run.df.summ.big.both.diag.2$n_seroconv_2018 [run.df.summ.big.both.diag.2$clusterID==234] # 6

run.df.summ.big.both.diag.2$medianAge2023 [run.df.summ.big.both.diag.2$clusterID==13] #53.6
run.df.summ.big.both.diag.2$medianAgeFARV [run.df.summ.big.both.diag.2$clusterID==13] #39
run.df.summ.big.both.diag.2$perc_MSM [run.df.summ.big.both.diag.2$clusterID==13] # 88
run.df.summ.big.both.diag.2$perc_HA_Van [run.df.summ.big.both.diag.2$clusterID==13] # 76.7
run.df.summ.big.both.diag.2$perc_HA_VanIsl [run.df.summ.big.both.diag.2$clusterID==13] #6
run.df.summ.big.both.diag.2$perc_PREP [run.df.summ.big.both.diag.2$clusterID==13] # 0
run.df.summ.big.both.diag.2$n_seroconv_2018 [run.df.summ.big.both.diag.2$clusterID==13] # 3

#cluster 33
clust.comp[clust.comp$clusterID==33,]
run.df.summ.big.both.diag.2$text_noprepsim [run.df.summ.big.both.diag.2$clusterID==33] # ""10.5 (9.5-11.4)"
run.df.summ.big.both.diag.2$text_prepsim [run.df.summ.big.both.diag.2$clusterID==33] #  ""4.6 (4.3-5)"
run.df.summ.big.both.diag.2$mean.diag.averted [run.df.summ.big.both.diag.2$clusterID==33] # "5.9
run.df.summ.big.both.diag.2$lower.diag.averted [run.df.summ.big.both.diag.2$clusterID==33] # "5.2
run.df.summ.big.both.diag.2$upper.diag.averted [run.df.summ.big.both.diag.2$clusterID==33] # "6.4

#245
clust.comp[clust.comp$clusterID==245,]
run.df.summ.big.both.diag.2$text_noprepsim [run.df.summ.big.both.diag.2$clusterID==245] # "22.6 (20.3-24.9)"
run.df.summ.big.both.diag.2$text_prepsim [run.df.summ.big.both.diag.2$clusterID==245] #  " "2.6 (2.4-2.8)"
run.df.summ.big.both.diag.2$mean.diag.averted [run.df.summ.big.both.diag.2$clusterID==245] # 20
run.df.summ.big.both.diag.2$lower.diag.averted [run.df.summ.big.both.diag.2$clusterID==245] #  17.9
run.df.summ.big.both.diag.2$upper.diag.averted [run.df.summ.big.both.diag.2$clusterID==245] # 22.1

```



## Join estimated and observed total diagnosed cases to calculate averted by cluster
```{r}
# merge observed (raw n) with predicted (no prep) cases
# run.df.summ.big2<-run.df.summ.big %>% left_join(clust.obs.cum, by=c("clusterID"))
# # colnames(run.df.summ.big2)
# 
# #make long
# run.df.summ.big2.long<-run.df.summ.big2 %>% pivot_longer(cols=c("n.sample","n.diag","total.observed"),
#                                                         names_to = "metric",
#                                                         values_to = "totaldiag")
# 
# #Merge in the simulated data fit to observed data (PrEP sim.)
# 
# #summarize total observed and predicted by cluster
# run.df.summ.big2.long %>%
#   filter(clusterID %in% med.active.clust) %>%
#   ggplot()+
#   geom_point(aes(x=as.factor(clusterID), group=metric, 
#                y=totaldiag, color=metric), 
#            position=position_dodge2(width=0.9),alpha=0.5,shape=18,size=1)+
#   geom_point(aes(x=as.factor(clusterID), group=metric, 
#                y=mean(totaldiag), color=metric), stat="identity",
#              position=position_dodge2(width=0.9),shape=17,size=3, alpha=0.7)+
#   pubTheme+
#   theme(legend.position = "top")+
#   labs(x=NULL,y="Total observed and predicted new diagnoses")
# ggsave(paste0(f.out,"TotalCases_ObservedPred_byCluster.png"),height=6,width=8.5)
# 
# #Mutate cases averted daily
# run.df.summ.big2<-run.df.summ.big2 %>% mutate(total.averted = n.diag - total.observed)
# 
# run.df.summ.big.avert<- run.df.summ.big2 %>% group_by(clusterID) %>%
#   dplyr::summarise(cases.averted=mean(total.averted))
# sum(run.df.summ.big.avert)
# run.df.summ.big.avert$clusterID==9999
# run.df.summ.big.avert[run.df.summ.big.avert$clusterID==9999,]
```

# MODELING FACTORS ASSOCIATED WITH CASES AVERTED

## Poisson/NegBin model of cases averted in clusters: assoc with median age, % HA, % population
```{r}
#association with age?
#cant have negatives for poisson
obs.expect.all<-run.df.summ.big.both.diag.2 %>% 
    filter(clusterID %in% active.clust)

#join seed data for 2017 size
seed.df$clusterID<-factor(seed.df$clusterID)
obs.expect.all<-left_join(obs.expect.all, seed.df, by="clusterID")

#change so refernce is pwid
obs.expect.all$risk<-str_replace_all(obs.expect.all$risk, "PWID","A_PWID")
# obs.expect.all$size2017
# head(obs.expect.all)

#Change negatives to zeroes
obs.expect.all$cases.averted.z<-obs.expect.all$mean.diag.averted
obs.expect.all$cases.averted.z[which(obs.expect.all$cases.averted.z<0)]<-0
# hist(obs.expect.all$cases.averted.z)

#instead of removing zeroes, normalize to the lowest value
obs.expect.all$cases.averted<-obs.expect.all$mean.diag.averted
min<--min(obs.expect.all$mean.diag.averted, na.rm = T)
obs.expect.all$cases.averted<-round(obs.expect.all$cases.averted+min+1)
# hist(obs.expect.all$cases.averted)

#distribution of values
hist(obs.expect.all$cases.averted)
mean(obs.expect.all$cases.averted)
var(obs.expect.all$cases.averted)
#negbinomial because overdispersed

#Join cluster composition data
# clust.comp$clusterID<-factor(clust.comp$clusterID)
# obs.expect.all<-obs.expect.all %>% left_join(clust.comp, by="clusterID")

#quick look
obs.expect.all %>%
  ggplot()+
  geom_point(aes(y=cases.averted, x=log(size2017),color=risk))
obs.expect.all %>%
  ggplot()+
  geom_point(aes(y=cases.averted, x=medianAge2023,color=risk))
```

## poisson
```{r}
#simple model
mod.null<-glm(data= obs.expect.all, cases.averted ~ NULL, family = poisson(link = "log"),)

mod0<-glm(data= obs.expect.all, cases.averted ~ log(size2017), family=poisson(link = "log"))
summary(mod0)
exp(confint(mod0))
anova(mod.null,mod0,test="LRT") #YES

mod<-glm(data= obs.expect.all, cases.averted ~ log(size2017) +perc_MSM, family=poisson(link = "log"))
summary(mod)
exp(mod$coefficients)
exp(confint(mod))
anova(mod0,mod,test="LRT") #YES add perc MSM

mod1<-glm(data= obs.expect.all, cases.averted ~ log(size2017)+perc_MSM+medianAge2023, family=poisson(link = "log"))
summary(mod1)
exp(mod$coefficients)
exp(confint(mod))
anova(mod,mod1,test="LRT") 

#add ha in there?
mod2<-glm(data= obs.expect.all, cases.averted ~ log(size2017)+perc_MSM+medianAge2023+perc_HA_Van, family=poisson(link = "log"))
summary(mod2)
exp(mod2$coefficients)
anova(mod1,mod2,test="LRT")  #No add HA van

#other HAs
mod2.1<-glm(data= obs.expect.all, cases.averted ~ log(size2017)+perc_MSM+medianAge2023 + perc_HA_VanIsl, family=poisson(link = "log"))
anova(mod2,mod2.1,test="LRT")  #NOPE
mod2.2<-glm(data= obs.expect.all, cases.averted ~ log(size2017)+perc_MSM+medianAge2023+ perc_HA_Fras,family=poisson(link = "log"))
anova(mod2,mod2.2,test="LRT")  #NOPE
mod2.3<-glm(data= obs.expect.all, cases.averted ~ log(size2017)+perc_MSM+medianAge2023 + perc_HA_North,family=poisson(link = "log"))
anova(mod2,mod2.3,test="LRT")  #NOPE

#add prep in there?
mod3<-glm(data= obs.expect.all, cases.averted ~ log(size2017)+perc_MSM+medianAge2023 + perc_PREP, family=poisson(link = "log"))
anova(mod2,mod3,test="LRT")  #No to adding prep
summary(mod3)
exp(mod3$coefficients)
exp(confint(mod3))

#EXport a table
finalmodel<-mod2
pois<-exp(finalmodel$coefficients) %>% as.data.frame()
pois$factor <-rownames(pois)
colnames(pois)[1]<-"Mean"
pois2<-exp(confint(finalmodel)) %>% as.data.frame()
pois2$factor <-rownames(pois2)
pois.mod<-left_join(pois, pois2)
pois.mod<-pois.mod[,c(2,1,3,4)]
pois.mod
#pvals
s<-summary(finalmodel)

write.csv(pois.mod, paste0(f.out, "pois.mod.caseaverted.csv"))

```

## Build negative binom
```{r}

#simple model
mod0<-glm.nb(data= obs.expect.all, cases.averted ~ size2017)
summary(mod0)
exp(confint(mod0))


mod1<-glm.nb(data= obs.expect.all, cases.averted ~ size2017+ risk)
summary(mod1)
exp(confint(mod1))

mod<-glm.nb(data= obs.expect.all, cases.averted ~ perc_MSM + medianAge2023)
summary(mod)
exp(mod$coefficients)
exp(confint(mod))

#add ha in there?
mod2<-glm.nb(data= obs.expect.all, cases.averted ~ perc_MSM + medianAge2023+perc_PREP)
summary(mod2)
exp(mod2$coefficients)

#add prep in there?
mod3<-glm.nb(data= obs.expect.all, cases.averted ~ perc_MSM + medianAge2023+perc_HA_Van+perc_PREP)
summary(mod3)
exp(mod3$coefficients)
exp(confint(mod3))

# Repeat with the zero-adj instead of up to zero

mod.z<-glm.nb(data= obs.expect.all, cases.averted.z ~ perc_MSM + medianAge2023)
summary(mod.z)
exp(mod.z$coefficients)
exp(confint(mod.z))

#add prep
mod2.z<-glm.nb(data= obs.expect.all, cases.averted.z ~ perc_MSM + medianAge2023+perc_PREP)
summary(mod2.z)

#add prep in there?
mod3.z<-glm.nb(data= obs.expect.all, cases.averted.z ~ perc_MSM + medianAge2023+perc_HA_Van+perc_PREP)
summary(mod3.z)
exp(mod3.z$coefficients)
exp(confint(mod3.z))
```

## Plot cases averted across clusters
```{r}
clusterorder<-obs.expect.all$clusterID[obs.expect.all$active==1]
obs.expect.all$clusterID<-factor(obs.expect.all$clusterID)

##PLOT THIS
obs.expect.all %>%
  filter(clusterID %in% med.active.clust) %>%
  ggplot()+
  geom_bar(aes(x=clusterID, y=cases.averted, fill=risk), stat="identity",alpha=0.7)+
  pubTheme+
  scale_fill_manual(values=riskcolz)+
  labs(x="Active clusters",y="Cases averted (observed - predicted stoch.)")+
  guides(fill=guide_legend(title="Predominant population"))+
  theme(legend.position = c(0.9,0.95), legend.justification = c(1,1),)+
  scale_y_continuous(limits=c(0,50))
        # axis.text.x=element_blank()
        
  # scale_y_continuous(breaks=seq(-15,80,5),expand=c(0.1,0))
ggsave(paste0(f.out,"/CasesAvertedByRiskGrp.png"),width=6,height=3.5)

```


## EXTRAS

## plots of outbreaks across clusters over time
```{r}
# run.df.all.big<-bind_rows(run.df.all)
# run.df.all.big<-run.df.all.big %>% mutate(date = infection.times + dayzero)
# 
# po<-run.df.all.big%>%
#   filter(clusterID %in% med.active.clust) %>%
#   ggplot()+
#   geom_bar(aes(x=as.Date(date), group=run, color=as.factor(clusterID)),alpha=0.2)+
#   # scale_color_viridis_c(direction = -1)+
#   pubTheme+
#   # scaleDateFlex+
#   labs(x="Date diagnosed", y="Density")+
#   theme(legend.position = "none")+
#   facet_wrap(~as.factor(clusterID),nrow=6)
#   # coord_cartesian(xlim=as.Date(c("2020-12-01","2021-07-01")))
# po
# ggsave(paste0(f.out,"Density_Sims_DiagOverTime_clusters.png"),height=4,width=4)

```

#extra plots
```{r}
#summarize total averted by cluster
# run.df.summ.big2 %>%
#   filter(clusterID %in% med.active.clust) %>%
#   ggplot()+
#   geom_violin(aes(x=as.factor(clusterID), group=as.factor(clusterID), y=n.diag))+
#   geom_col(aes(x=as.factor(clusterID), group=as.factor(clusterID),
#                y=total.observed),alpha=0.2)+
#   pubTheme+
#   # theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
#   # theme(legend.position = "none")+
#   labs(x=NULL,y="Total averted diagnoses")
# ggsave(paste0(f.out,"BoxP_Dailycases_Averted.png"),height=6,width=7)

# 
# 
# 
# #Repeat but no stack, just a line and ribbon overlaid
# y.lim2<-c(0,max(can.var.cases.red$mean.dailycase))
# p.cases.obs.pred.rib<-ggplot()+
#     annotate(geom="rect",xmin = as.Date(int.start), xmax = as.Date(int.end),
#              ymin = y.lim2[1], ymax =y.lim2[2],
#              color="grey85",size=0,  fill="grey80")+
#     annotate(geom="text",x =( as.Date(int.start)+int.duration/2),
#              y=(y.lim2[2]/100)*99, vjust=1, hjust=0.5,
#              label=int.descrip,fontface="italic",color="grey25",size=3)+
#     scale_fill_manual(values=pred.pal)+
#     pubThemeDate+
#     geom_bar(data=can.var.cases.red,
#              aes(x=date, y=mean.dailycase, group=type), stat="identity",width=1)+
#   geom_line(data=cases.mean.date, aes(x=date,y=mean.dailycase.diagnosed),alpha=0.3)+
#   geom_ribbon(data=cases.mean.date, aes(x=date,ymin=lower.dailycase.diagnosed,
#                                         ymax=upper.dailycase.diagnosed),alpha=0.3)+
#     scaleDateFlexMoreButLess2+
#     scale_y_continuous(expand=c(0,0))+
#     # GlobFillScale+
#     theme(legend.position = "top")+
#     labs(x=NULL, y=paste0("Average daily diagnosed ",Focal.var," cases"))+
#     coord_cartesian(ylim=y.lim2)
#   
# p.cases.obs.pred.rib
# ggsave(paste0(f.out,"canada.var.cases.pred.obs_rib.png"),height=4,width=4,unit="in")
# 
# ## grobbed plot with overlaid possible simulations and then case incidence
# plot_grid(p.sims, p.cases.obs.pred, labels=LETTERS[1:2], align="h",nrow=1)
# ggsave(paste0(f.out,"Sims_ObsPredCases.png"), height=3.5,width=8)
```


## Export summarized simulation draws as text
```{r}
# #overall
# 
# #sampled cumulative 
# n.cum<-c()
# #max daily sampled
# n.max.daily<-c()
# #mean sampled outbreak size
# mean.size<-c()
# #max sampled outbreak size
# max.size.s<-c()
# 
# #diagnosed cum
# n.cum.diag<-c()
# 
# #max daily diagnosed, etc etc
# n.max.daily.diag<-c()
# mean.size.diag<-c()
# max.size.diag<-c()
# 
# #all infected cum (ew)
# n.cum.all<-c()
# n.max.daily.all<-c()
# mean.size.all<-c()
# max.size.all<-c()
# 
# 
# #for each draw
# for (i in 1:length(draw.list)){
#   
#   draw.list[[i]]$date<-factor(draw.list[[i]]$date)
# 
#   temp<-draw.list[[i]] %>% dplyr::group_by(run) %>%
#     dplyr::summarize(intro.date=intro.date[1],
#                      n.all=n(),
#                      n.diag=length(which(sampled==1)),
#                      n.sample=length(which(sampled==1)),
#                      max.time=max(infection.times))
#   
#   intro.n<-nrow(temp)
#   per1.n<-length(which(temp$intro.date <= per1[2]))
#   per2.n<-intro.n-per1.n
#   sing<-length(which(temp$n.sample==1))
#   max.l<-max(temp$max.time)
#   median.l<-median(temp$max.time)
# 
#   #SSAMPLES
#   temp2<-draw.list[[i]] %>% dplyr::group_by(date) %>% 
#     dplyr::summarise(dailycase=sum(sampled))
#   
#   cum.s<-sum(temp2$dailycase)
#   max.day.s<-max(temp2$dailycase)
#   max.s<-max(temp$n.sample)
#   mean.s<-mean(temp$n.sample)
#   
#   ## add to running vectors
#   n.sing<-c(n.sing, sing)
#   n.intro.per1<-c(n.intro.per1,per1.n)
#   n.intro.per2<-c(n.intro.per2, per2.n)
#   max.longs<-c(max.longs,max.l)
#   med.longs<-c(med.longs,median.l)
#   
#   n.cum<-c(n.cum, cum.s)
#   n.max.daily<-c(n.max.daily,max.day.s)
#   mean.size<-c(mean.size,mean.s)
#   max.size.s<-c(max.size.s,max.s)
#   
#   #repeat for all - these are currently same as diagnosed b/c undiagnosed removed for mem
#   temp3<-draw.list[[i]] %>% dplyr::group_by(date) %>% 
#     dplyr::summarise(dailycase=n())
#   
#   cum.all<-sum(temp3$dailycase)
#   max.day.all<-max(temp3$dailycase)
#   max.s.all<-max(temp$n.all)
#   mean.s.all<-mean(temp$n.all)
#   
#   ## add to running vectors
#   n.cum.all<-c(n.cum.all, cum.all)
#   n.max.daily.all<-c(n.max.daily.all,max.day.all)
#   mean.size.all<-c(mean.size.all,mean.s.all)
#   max.size.all<-c(max.size.all,max.s.all)
#   
#   #repeat for diagnosed
#   temp4<-draw.list[[i]] %>% dplyr::group_by(date) %>% 
#     dplyr::summarise(dailycase=sum(diagnosed))
#   
#   cum.diag<-sum(temp4$dailycase)
#   max.day.diag<-max(temp4$dailycase)
#   max.s.diag<-max(temp$n.diag)
#   mean.s.diag<-mean(temp$n.diag)
#   
#   #Day of max daily diag
#   d.max.daily.diag<-as.Date(temp4$date[which(temp4$dailycase==max.day.diag)][1])
#     
#   #by period
#   temp4.p1<-draw.list.summ[[i]] %>% filter(intro.date<=per1[2])
#   cum.diag.p1<-sum(temp4.p1$n.diag,na.rm = T)
#   temp4.p2<-draw.list.summ[[i]] %>% filter(intro.date>=per2[1])
#   cum.diag.p2<-sum(temp4.p2$n.diag,na.rm = T)
#   
#   
#   ## add to running vectors
#   n.cum.diag<-c(n.cum.diag, cum.diag)
#   n.max.daily.diag<-c(n.max.daily.diag,max.day.diag)
#   mean.size.diag<-c(mean.size.diag,mean.s.diag)
#   max.size.diag<-c(max.size.diag,max.s.diag)
#   
#   n.cum.diag.p1<-c(n.cum.diag.p1, cum.diag.p1)
#   n.cum.diag.p2<-c(n.cum.diag.p2, cum.diag.p2)
#   day.max.daily.diag<-c(day.max.daily.diag,  d.max.daily.diag)
# }
# 
# ## add all of these in the text export
# cat(paste0("SIMULATIONS overview",
#             "\nMean # singletons sampled: ",mean.95ci.X(n.sing,0), 
#             "\nMean intros period 1: ",mean.95ci.X(n.intro.per1,0),
#             "\nMean intros period 2: ",mean.95ci.X(n.intro.per2,0), 
#             "\nMax longevity: ",mean.95ci.X(max.longs,0), 
#             "\nMedian longevity: ",mean.95ci.X(med.longs,0), 
#            
#             "\nMean cumulative sampled cases: ",mean.95ci.X(n.cum,0), 
#             "\nMax daily sampled cases: ",mean.95ci.X(n.max.daily,0), 
#             "\nMean intro sampled size: ",mean.95ci.X(mean.size,0), 
#             "\nMax intro sampled size: ",mean.95ci.X(max.size.s,0), 
# 
#             "\nMean cumulative cases diag: ",mean.95ci.X(n.cum.diag,0), 
#             "\nMean cumulative cases diag period 1: ",mean.95ci.X(n.cum.diag.p1,0), 
#             "\nMean cumulative cases diag period 2: ",mean.95ci.X(n.cum.diag.p2,0), 
#             "\nMax daily cases diag: ",mean.95ci.X(n.max.daily.diag,0), 
#             "\nMean intro size diag: ",mean.95ci.X(mean.size.diag,0), 
#             "\nMax intro size diag: ",mean.95ci.X(max.size.diag,0), 
#                
#             "\nMean cumulative cases all: ",mean.95ci.X(n.cum.all,0), 
#             "\nMax daily cases all: ",mean.95ci.X(n.max.daily.all,0), 
#             "\nMean intro size all: ",mean.95ci.X(mean.size.all,0), 
#             "\nMax intro size all: ",mean.95ci.X(max.size.all,0)), 
#     append=T, sep="\n", file=text.out)
# 
# ## Export for delay of peak incidence
# can.var.max.cases<-max(can.var.cases$avgVOC_cases,na.rm = T)
# day.can.var.max<-can.var.cases$date[which(can.var.cases$avgVOC_cases == can.var.max.cases)][1]
# cat(paste0("\nNumber days earlier incidence peak: ", 
#            mean.95ci.X(as.numeric(day.can.var.max - day.max.daily.diag))),
#      append=T, sep="\n", file=text.out)
# 
# 
# #Add export for averted cases diagnosed per day
# print(paste0("\nMean cases diagnosed averted per day: ", mean.95ci.X(n.cum.diag/int.duration,1),
#            "\nMean cases diagnosed averted per day period 1: ", mean.95ci.X(n.cum.diag.p1/(as.numeric(diff(per1)+1)),1),
#      "\nMean cases diagnosed averted per day period 2: ", mean.95ci.X(n.cum.diag.p2/(as.numeric(diff(per2)+1)),1)))
# 
# ## Export percent averted
# 
# # % additional maximum daily incidence
# can.var.max.cases<-max(can.var.cases$avgVOC_cases,na.rm = T)
# max.cases.averted.perc<-n.max.daily.diag/can.var.max.cases*100
# maxCI.averted<-mean.95ci.X(max.cases.averted.perc)
#  
# # percent additional cases cumulative (considering sampled only)
# max.samp.cases.averted.perc<-(n.max.daily/can.var.max.cases)*100
# maxCI.samp.averted<-mean.95ci.X(max.samp.cases.averted.perc)
# 
# 
# ## % Additional cumulative cases 
# 
# # can.var.total.cases
# mean.cases.averted.perc<-(n.cum.diag/can.var.total.cases)*100
# #percent additional cases cumulative
# meanCI.averted<-mean.95ci.X(mean.cases.averted.perc)
# 
# 
# #percent additional cases cumulative (considering sampled only)
# mean.samp.cases.averted.perc<-(n.cum/can.var.total.cases)*100


```


## mean and 95% CI daily cases across runs for each cluster
```{r}
# has to be a factor to group by, count up occurrences
# draw.list.all$date<-factor(draw.list.all$date)
# 
# #in each draw (gruoping multiple runs per draw), sum of daily cases
# cases.run.date<-draw.list.all %>%
#   group_by(draw, date) %>%
#   summarize(dailycase.sampled=sum(sampled),
#             dailycase.diagnosed=sum(diagnosed),
#             dailycase.all=n())
# 
# # head(cases.run.date,n=100) 
# # sum(cases.run.date$dailycase[cases.run.date$draw==10])
# 
# #plot each one as trajectory of daily cases sampled
# ggplot(cases.run.date)+
#   geom_line(aes(x=as.Date(date), y=dailycase.sampled, group=draw,color=draw),stat="identity",alpha=0.3)+
#   pubThemeDate+
#   theme(legend.position = "none")+
#   labs(x="Date",y="Daily cases averted (sampled)")
# ggsave(paste0(f.out,"DailyCasesAvertedSampled.AcrossDraws.png"),height=4,width=5 )
# 
# #plot each one as trajectory of daily cases diagnosed
# ymax<-max(cases.run.date$dailycase.diagnosed)*1.2
# p.sims<-ggplot(cases.run.date)+
#   geom_line(aes(x=as.Date(date), y=dailycase.diagnosed, group=draw,color=draw),stat="identity",alpha=0.4)+
#   pubThemeDate+
#   scaleDateFlex+
#   scale_color_viridis_c()+
#   theme(legend.position = "none")+
#   labs(x=NULL,y="Daily cases averted (diagnosed)")+
#   scale_y_continuous(expand=c(0,0))+
#   coord_cartesian(ylim=c(0,ymax))
# p.sims
# ggsave(paste0(f.out,"DailyCasesAvertedDiagnosed.AcrossDraws.png"),height=4,width=5 )
# 
# #check cumsum (what is the sum of cum?) of each one
# # cases.run.date %>% group_by(draw) %>% summarize(total=sum(n))
# 
# #now summarize daily cases across runs
# cases.mean.date<-cases.run.date %>% 
#   dplyr::group_by(date) %>%
#   dplyr::summarise(mean.dailycase.sampled=mean(dailycase.sampled),
#                    sd.dailycase.sampled=sd(dailycase.sampled),
#                    
#                    mean.dailycase.diagnosed=mean(dailycase.diagnosed),
#                    sd.dailycase.diagnosed=sd(dailycase.diagnosed),
#                    
#                    mean.dailycase.all=mean(dailycase.all),
#                    sd.dailycase.all=sd(dailycase.all))
# 
# #mutate confints - shoulld write a function for this in future
# cases.mean.date<-cases.mean.date %>% mutate(margin.err.sampled= qt(p=0.05/2, df=(n.draw-1),lower.tail=F)*sd.dailycase.sampled/sqrt(n.draw),
#                                             upper.dailycase.sampled=mean.dailycase.sampled + margin.err.sampled,
#                                             lower.dailycase.sampled=mean.dailycase.sampled - margin.err.sampled,
#                                             
#                                             margin.err.diagnosed= qt(p=0.05/2, df=(n.draw-1),lower.tail=F)*sd.dailycase.diagnosed/sqrt(n.draw),
#                                             upper.dailycase.diagnosed=mean.dailycase.diagnosed + margin.err.diagnosed,
#                                             lower.dailycase.diagnosed=mean.dailycase.diagnosed - margin.err.diagnosed,
#                                             
#                                             margin.err.all= qt(p=0.05/2, df=(n.draw-1),lower.tail=F)*sd.dailycase.all/sqrt(n.draw),
#                                             upper.dailycase.all=mean.dailycase.all + margin.err.all,
#                                             lower.dailycase.all=mean.dailycase.all - margin.err.all)
# # head(cases.mean.date)
# 
# #plot the mean daily cases over time
# px1<-ggplot(cases.mean.date)+
#   geom_line(aes(x=as.Date(date), y=mean.dailycase.sampled),stat="identity")+
#   geom_ribbon(aes(x=as.Date(date), ymin=lower.dailycase.sampled, ymax=upper.dailycase.sampled),alpha=0.5)+
#   pubThemeDate+
#   labs(y="Mean daily cases averted (sampled)",x="Date")
# px1
# ggsave(paste0(f.out,"meanCI_DailyCasesAverted_Sampled.png"),width = 3,height=3)
# 
# #plot the mean daily cases diagnosed over time
# px2<-ggplot(cases.mean.date)+
#   geom_line(aes(x=as.Date(date), y=mean.dailycase.diagnosed),stat="identity")+
#   geom_ribbon(aes(x=as.Date(date), ymin=lower.dailycase.diagnosed, ymax=upper.dailycase.diagnosed),alpha=0.5)+
#   pubThemeDate+
#   labs(y="Mean daily cases averted (diagnosed)",x="Date")
# px2
# ggsave(paste0(f.out,"meanCI_DailyCasesAverted_Diagnosed.png"),width = 3,height=3)
# 
# #grob it baby
# plot_grid(px1, px2, labels=LETTERS[1:2], nrow=1, align="h")
# ggsave(paste0(f.out,"meanCI_DailyCasesAverted_Sampled-Diagnosed.png"),width = 7,height=3.6)
# 
# #what is the cumsum of dailycases
# sum(cases.mean.date$mean.dailycase.sampled)
# mean.95ci.X(n.cum,0)
# 
# sum(cases.mean.date$mean.dailycase.diagnosed)
# 
# #export 
# write.csv(cases.mean.date,file = paste0(f.out,"casesAverted.MeanDaily.csv"))
```

## check p.seqperdiag
```{r}
dtp.full<-read.csv("../../00_data/01_clean/clean_output/dtp.csv")
dtp.phy<-read.csv("../../00_data/01_clean/clean_output/phyloDTP.csv")

dtp.full.count<-dtp.full %>% 
  filter(PREV_ARV != TRUE) %>%
  filter(as.Date(FARVDT) >= as.Date("2018-01-01")) %>%
  dplyr::summarise(total=n())
dtp.phy.count<-dtp.phy %>% 
  filter(PREV_ARV != TRUE) %>%
  filter(as.Date(FARVDT) >= as.Date("2018-01-01")) %>%
  dplyr::summarise(total=n())


p.seqpercase
dtp.phy.count/dtp.full.count #95%
```

